EEG Waveform Analysis with applications to Brain Computer Interfaces

### 2018-04-30 sensors.tex
Changes up

% Abstract (Do not use inserted blank lines, i.e. \\) 
\abstract{The Electroencephalography (EEG) is not just a mere clinical tool anymore.  It has become the de-facto mobile, portable, non-invasive brain imaging sensor to harness brain information in real time and translating or decoding brain signals that can be used to diagnose disease or implement Human Computer Interaction devices.  The automatic approach which is based on using digital tools to detect the cloaked information buried in the signal, outshines the research done by the clinical EEG community which was based intensively on EEG waveforms and the structure of signal plots.  The purpose of this work is to help to start to fill this gap by doing a review and description of the automatic methods that have been used to detect patterns in the waveforms, and to perform a benchmarking analysis to determine different characteristics of those methods that aim to detect specific waveforms mimicking what physicians has been doing since the inception of this fruitful technology. }

@@

Current society is demanding technology to once and for all provide the means to realize the utopia of social inclusion for people with disabilities.  Additionally, a worldwide aging population is also claiming for solutions to extend active lifestyles throughout all life, including additional years that medicine is allowing humanity to have \citep{Lutz2008}.
At the same time,  the digital gadgets and the digital revolution have modified the way people interact with all their devices.  A new emerging digital society is consolidating and smart wearable sensors are starting to be used routinely to obtain bio-signals, in an active or passive way (Wolpaw Communication and control)
Although this communication is based on muscular movement \citep{Guger2017}, all these trends are precisely pushing this boundary beyond the confines of the body and beyond the limitation of human movement.  A new form of Human Machine communication which directly connects the Central Nervous System to a machine is currently being under development: Brain Machine Interfaces.  Moreover, as long as computers are being used inside every machine that humanity posses \citep{Domingo2012} this discipline is specializing as Brain Computer Interfaces (BCI) or Brain-Neural Computer Interfaces (BNCI).
On the center of all this hype, we can find a hundredth year old technology, rock-solid as a diagnosis tool, which greatly benefited from the shrinkage of sensors, the increase in computer power and the widespread development of wireless protocols and advanced electronics: the Electroencephalogram (EEG) \citep{Schomer2010}.

EEG sensors are wearable \citep{Puce2017} non-invasive, portable and mobile \citep{DeVos2014}, with excellent temporal resolution, and acceptable spatial resolution \citep{Hartman2005}.  This humble diagnosis device has been transformed into currently the best approach that we posses to precisely detect, out-of-the lab in an ambulatory context, information from the Central Nervous System and to use that information to volitionally drive our cars, steer our drones, write our emails, and control our wheelchairs \citep{Yuste2017}.

The clinical and historical approach to analyze EEG signals were based on detecting visual patterns out of the EEG trace or polygraph (Atlas of EEG patterns): multichannel signals were extracted and plotted over a piece of paper, continuously. Electroencephalographers or Electroencephalography technician have decoded and detected patterns along the signals by visually inspecting them \citep{Schomer2010}.  Even nowadays clinical EEG remains a visually interpreted test \citep{Hartman2005}.

Automatic processing, or quantitave EEG, was based first on analog electronic devices and later on computerized digital processing methods \citep{Jansen1991}.  They implemented mathematically and algorithmically complex procedures to decode the information with outstanding success \citep{Yuste2017}.  The best materialization of the automatic processing of EEG signals lays in the BCI discipline, where $71.2\%$ is based on EEG \citep{Guger2017}.

@@

Hence, the traditional and knowledgeable approach was mainly overshadow particularly in BCI Research, and the waveform of the EEG was replaced by pragmatically sound procedures that were difficult to link to existing clinical EEG knowledge.  We aim to help fix this gap by providing a review of the methods which emphasize the waveform, the shape of the EEG signal and that can help to decode them in an automated way.

The aim of this study is twofold: first to review current literature of EEG processing techniques which are based on analysis of the waveform.  The second is to evaluate and study these methods by analyzing its classification performance against a pseudo-real dataset. We aim to outline a better understanding of the different approaches.  We believe that the importance of waveform analysis methods, as described here, is that by using this methodology, collaboration could be fostered because there is a clear description and characterization of the signal, where the extensive literature which explores clinical EEG can be reviewed from the same shared perspective \citep{Nijboer2009,Wei2017}.

@@

The Electroencephalogram is one of the most widespread used device to capture brain signals.  It was initially developed by Hans Berger in 1924 and has been extensively used for decades to diagnose neural diseases and other medical conditions.

The first characterization that Dr. Berger detected was the visual cortical alpha wave, the \textit{Berger Rythm} \citep{Jansen1991}.  He understood that the amplitude and shape of this rhythm was coherently associated to a cognitive action (eyes closing).  
We should ask ourselves if the EEG research explosion that came after that discovery would have happened, if it weren't so evident that the shape alteration was due to a very simple and verifiable cognitive process.

The EEG signal is highly complex.  It can be modeled as a linear stochastic process with great similarities to noise \citep{Thakor2004}.  They are measured in microvolts, and those slightly variations are contaminated with heavy endogenous artifacts and exogenous spurious signals.

\begin{figure}[H]
\centering
\includegraphics[width=6cm]{eegsample.png}
\caption{Sample EEG signal captured and plotted on a 2D image using the popular and commercial EPOC Emotiv using EEGLAB.}
\label{fig:sampleeeg}
\end{figure}

The device that captures these small variations in the current potentials over the scalp is called the Electroencephalograph.  An  explanation on the particularities and modelling of EEG can be obtained from \citep{Jackson2014}, as well as in the description of the electrophysiological aspects of the method \citep{Haberman2012}.

EEG signals have the following general components:

\begin{itemize}
\item Artefacts: endogeneous exogeneous as well as physiological or non-physiological (EEG MEG REview).
\item Non-Stationarity: the statistical parameters that describe the EEG as a random process are not conserved through time, i.e. its mean and variance, and any other higher-order moments are not time-invariant \citep{Jansen1991}.
\item DC drift and trending
\item Basal EEG activity
\item Inter-subject and intra-subject variability:It is known that EEG can be affected by the person behavior like sleep hygiene, caffeine intake, smoking habit, alcohol intake before EEG \citep{Farzan2017}.
\end{itemize}

Overall, EEG can be characterized by their phase, amplitude,  frequency and \textit{waveform}.

\begin{itemize}
\item Spontaneous: generally treated as noise or basal EEG.
\item Evoked: activity that is time-locked to an incoming stimulus or an executed motor action.  In contrast it is often called Induced activty
\end{itemize}

Understood according to the belongings or not to a repeated rhythm.

\begin{itemize}
\item Rhythmic. EEG activity consisting in waves of approximately constant frequency.  It is often abbreviated RA (regular rythmic activity).
\item Arrhythmic. EEG activity in which no stable rhythms are present.
\item Dysrhythmic. Rhythms and/or patterns of EEG activity that characteristically appear in patient groups or rarely or seen in healthy subjects.
\end{itemize}

Additionally, the number of electrodes and their positions over the scalp determines a \textbf{Spatial Structure}: waveforms can be generalized, focal or lateralized (EEG).

Finally, indexes can be derived as  CFC, Cross Frequency Coupling,  Phase-Amplitude Coupling, Phase-Phase Coupling.

The traditional clinical approach consisted in analyzing the paper strip that was generated by the plot of the signal obtained from the device.  Expert technician and physicians analyzed visually the plots looking for specific patterns that may give a hint of the underlying cognitive process or pathology.   Atlases and guidelines were created in order to help in the recognition of these complex patterns.   Even Video-electroencephalography scalp recordings are routinely used as a diagnostic tools \citep{Giagante2003} .  The EEG research has focused on temporal waveforms, and a whole branch of electrophenomenology has arisen around EEG \textit{graphoelements} \citep{Schomer2010}.

Sleep Research has been studied in this way by performing Polysomnographic recordings (PSG)  \citep{Rodenbeck2006}, where the different sleep stages are evaluated by visually marking waveforms or graphoelements in long-running electroencephalographic recordings, looking for patterns based on standardized guidelines.   Visual characterization includes the identification or classification of certain waveform components, or transient events, based on a subjective characterization (e.g. positive or negative peak polarity) or the location within the strip.  It is regular to establish an amplitude difference between different waveforms from which a relation between them is established and structured index are created (e.g. sleep K-Complex is well characterized based on rates between positive vs negative amplitude) (REFERENCE).  Other relevant EEG patterns for sleep stage scoring are alpha, theta, and delta waves,  sleep spindles, polysplindles, K-complexes, vertex sharp waves (VSW), and sawtooth waves (REM Sleep).

Moreover, in Epilepsy EEG data acquisition is crucial during the assessment of patients with focal epilepsies for potential seizure surgery, where the origin of the seizure activity must be reliably detected. The determination of the Epileptic Seizure onset is defined as the first electrical change seen in the EEG rhythm as compared to baseline or any clinical sign indicating seizure onset (Clinical EEG).   Most epilepsy patients also show characteristic interictal (or between-seizure) epileptiform discharges (IEDs) termed spike (<70 microsec duration), spike and wave, or sharp-wave (70–200 microsec duration) discharges. (Electroencephalography Introduction Canadian Epileptic Assosiation).  All these characterization are strictly based on the waveform shape.

@@

Finally, there are specific EEG patterns that can be used to determine Depth of anestisia and aEEG, amplitude integrated electroencephalography or cerebral function monitoring (REFERENCE) is used to determine coma levels based on shapes obtained from shapes observed in the EEG.

@@

The shape of the signal, the waveform, can be defined as the graphed line that represents the signal's amplitude plotted against time. It can  also called EEG biomarker,  EEG pattern, signal shape, signal form and a morphological signal (CITAS DONDE SE LAS LLAMA ASI).

Waveform characterization pays importance to context, both in a spatial and in a temporal sense (Jensen).  On the basis of contextual consideration, some specific waveform can be considered as noise while in other context is precisely the element which has a cognitive functional implication.

Shape or waveform analysis methods are considered as nonparametric methods (in opposition to statistical or dynamical models).  They can explore the amplitude, energy or more complex like the teager energy operand and L-Z Lempel-Ziv complexity measurement (thakor qEEG).

Slight variations in weather slopes or other features are scored may be important to certain applications.

It has
\begin{itemize}
\item characteristic shape, or waveform.
\item rising phase
\item falling phase
\item pronounced plateau
\item ripples and wiggles. 
\end{itemize}

\begin{itemize}
\item Amplitude
\item Arch
\item Frequency
\item Phase
\item Nonsinusoidal sinusoidal
\item Oscilation
\item Sawtooth: motor cortical beta oscillations
\item Sharpness
\item Spike-wave discharge
\item Transient event
\end{itemize}

Other characterization include, subjective definitions of sharper, arch comb or wicket shape, sawtooth, rectangular, spike-wave like, decay phase, voltage rise, peaks and troughs, short term voltage change around each extrema in the raw trace, peak and trough sharpness ratio, symmetry between rise and decay phase, slope ratio (steepness of the rise period to that of the adjacent decay period.  For instance,  descriptions like "Central trough is sharper and more negative that the adjacent troughs" are common in the literature.

\begin{itemize}
\item Attenuation (synonyms: suppression, depression). Reduction of amplitude of EEG activity resulting from decreased voltage. When activity is attenuated by stimulation, it is said to have been "blocked" or to show "blocking".
\item Hypersynchrony. Seen as an increase in voltage and regularity of rhythmic activity, or within the alpha, beta, or theta range. The term implies an increase in the number of neural elements contributing to the rhythm. (Note: term is used in interpretative sense but as a descriptor of change in the EEG).
\item Paroxysmal. Activity that emerges from background with a rapid onset, reaching (usually) quite high voltage and ending with an abrupt return to lower voltage activity. Though the term does not directly imply abnormality, much abnormal activity is paroxysmal.
\end{itemize}

\begin{itemize}
\item Monomorphic: Distinct EEG activity appearing to be composed of one dominant activity
\item Polymorphic: distinct EEG activity composed of multiple frequencies that combine to form a complex waveform.
\item Sinusoidal. Waves resembling sine waves. Monomorphic activity usually is sinusoidal.
\item Transient. An isolated wave or pattern that is distinctly different from background activity.
\end{itemize}

@@

%To verify the validity of the proposed framework and method, the public dataset 008-2014  \citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by  IRCCS Fondazione Santa Lucia, was used to perform a binary classification task on the provided signals.  The algorithm was implemented using the VLFeat  \citep{Vedaldi2010} Computer Vision libraries on MATLAB 2014a (Mathworks Inc., Natick, MA, USA).

In order to determine which processing method to be considered as a  "waveform template matching" we restrict ourselves to the following criteria:

\begin{enumerate}
\item The pattern can be identified and verified by visual inspection.
\item The pattern matching is performed in time-domain.
\item The analysis take account the shape of the plot of the signal.
\item The feature extraction procedure allows to create a Template database.
\item A Single Channel Processing approach.
\end{enumerate}

As described in (Mixed Domain Signal Analysis) the Pattern Matching problem in Signal processing is finding a signal given the region that best describes the structure of the pattern.

\subsection{Selected Algorithms}

@@

\begin{itemize}
\item Matching Pursuit
\item Permutation Entropy
\item Slope Horizontal Chain Code
\item Histogram of Gradient Orientations
\item Merging of Increasing and Decreasing Sequences
\end{itemize}

Although the term morphology has been used to identify this approach, We specifically excluded Mathematical Morphological based methods \cite{Yamaguchi2009}.  On the other hand, Morphological Component Analysis is a variant form of Blind Source Separation which can be considered the extension of Matching Pursuit algorithm. (cita a esa parte del paper).

\subsection{Excluded Methods}

\subsubsection{Waveform features}

One of the earliest approach to automatically process EEG data is the Peak Picking method.  Although of limited usability, peak picking has been used to determine latency of transient events in EEG \citep{Jaskowski2000,Zhang2011}.  Straightforward in its implementation, it consists in selecting a component, particularly a simple component based on the expected location of its more prominent deflection \citep{Ouyang2017}.  Particularly common in ERP Research, the name of many of the EEG features directly reference a peak within the component, e.g. P300 or P3a P3b or N100.  This leads to a natural way to classify them visually by selecting appropriate peaks and matching their positions and amplitudes in an orderly manner.  The letter provides the polarity (Positive or Negative) and the numbering shows the time referencing the stimulus onset, or the ordinal position of each peak (first, second, etc).

A related method is used in \citep{Alvarado-Gonzalez2016} where the area under the curve of the EEG is sumarized to derive a feature.  This was even used in the seminal work of Farwell and Donchin on P300 \citep{Farwell1988,WolpawJonathanR2012}. Additionally, a logarithmic graph of the peak-to-peak amplitude or aEEG \citep{Shah2015} is used nowadays in Neonatal Intensive Care Units.

Other works on EEG explored the idea to extend human capacities analyzing EEG waveforms  (automatic detection of k complexes), (A Waveform Analyzer Applied to the Human EEG) where a feature from amplitude and frequency of its signal and its derivative in time-domain is used.

Althought CASENET REFERENCe explored "waveform" structure they were purely based on spike detection based on feeding artificial neural networks.

 (fujimori, PAA, etc).  Uchida 1999

\subsubsection{Averaging Methods}

The methods that allow to identified waveforms are used to determine different alignments while averaging epochs.

This long-standing problem has been tackled from different perspectives.  Woody's Template Matching is perhaps the best effort as well as Pham, and Tuan and others EML.

Dictionary, template based method.  Many do not consider the waveform, albeit they do use a set of templates obtained from a dictionary.

Cross Covariance with Template or Cross Correlation or ACF: Autocorrelation Function: PHAM METHOD

Applying a FIR filter with templates

Krusienski et al 2007
Serby et al 2005

extended to wavelent analysis.

Dynamic Time Warping and other Warp Averaging methods were also used.

Wrapping Analysis

Warp averaging

\subsection{Matching Pursuit}

Pursuit algorithms refer, in their many variants, as blind source separation techniques that assume that the EEG signal is a linear combination of different sources that comes from template dictionaries.

We include them here because they are traditional in terms for the traditional view of Morphological Analysis or Shape domain analysis, though they do not restrict to our own definition of "shape domain processing".

However in terms of

\subsection{Permutation Entropy}

Bond and Pompe PE method.

This method is the best representative of Waveform Complexity: Lempel-Ziv method L-Z complexity are other variants which use a different definition.

Permutation Entropy a new feature for brain computer interfaces.

\subsection{Slope Horizontal Chain Code}

contour representation based on an adapted version of the Slope Chain Code (SCC)

This method can be encompass as a syntactic analysis technique, where the EEG segment is represented as a series of elementary patterns, similar to tokens.

\subsection{Histogram of Gradients}

\subsection{Merger of Increasing and Decreasing Sequences}

Stepwise downsampling

@@

Results are shown in Table 1 and in Figure 1.  Table 1 shows the performance to identify each letter of the standard P300 Speller Matrix.   Figure 2 shows how the performance increase when averaging is implemented.

>>>>

% Abstract (Do not use inserted blank lines, i.e. \\) 
\abstract{The Electroencephalography (EEG) is not just a mere clinical tool anymore.  It has become the de-facto mobile, portable, non-invasive brain imaging sensor to harness brain information in real time and translating or decoding brain signals that can be used to diagnose disease or implement Human Computer Interaction devices.  The automatic processing approach which is based on using quantitative algorithms to detect the cloaked information buried in the signal, outshines the research done by the clinical EEG community which was based intensively on EEG waveforms and the structure of signal plots.  Hence, the purpose of this work is to help to start to fill this gap by doing a review and description of the procedures that have been used to detect patterns in the waveforms, and to perform a benchmarking analysis of them.  }

% We aim to detect specific waveforms mimicking what physicians has been doing since the inception of this fruitful technology.

@@

%People with disabilities needs technological solutions for inclusion

%Old people is more likely to be disabled.

%The society is getting old.

%We also have digital gadgets everywhere, so the society needs new ways to interact with those devices.

%So far the interaction with devices has been based on muscular movement, but these trends are pushing the boundary beyond the confines of the body.

Current society is demanding technology to provide the means to realize the utopia of social inclusion for people with disabilities\citep{Wolpaw2002}.  Additionally, as societies are aging \citep{Lutz2008} the incidence of neuromuscular atrophies, strokes and other invalidating diseases is increasing.  Concurrently, the digital revolution and the pervasiveness of digital gadgets have modified the way people interact with the environment through these devices \citep{Domingo2012}.  All this communication is based on muscular movement \citep{Guger2017}, but these trends are pushing this boundary beyond the confines of the body and beyond the limitation of human movement.  A new form of Human Machine communication which directly connects the Central Nervous System (CNS) to a machine or computer device is currently being developed: Brain Machine Interfaces (BMI), Brain Computer Interfaces (BCI) or Brain-Neural Computer Interfaces (BNCI).

At the center of all this hype, we can find a hundredth year old technology, rock-solid as a diagnosis tool, which greatly benefited from the shrinkage of sensors, the increase in computer power and the widespread development of wireless protocols and advanced electronics: the Electroencephalogram (EEG) \citep{Schomer2010}.

EEG sensors are wearable \citep{Puce2017} non-invasive, portable and mobile \citep{DeVos2014}, with excellent temporal resolution, and acceptable spatial resolution \citep{Hartman2005}.  This humble diagnosis device is been transformed into currently the best approach to detect, out-of-the lab in an ambulatory context, information from the Central Nervous System and to use that information to volitionally drive cars, steer drones, write emails, or control wheelchairs \citep{Yuste2017}.

The clinical and historical tactic to analyze EEG signals were based on detecting visual patterns out of the EEG trace or polygraph\citep{Hartman2005}: multichannel signals were extracted and continuously plotted over a piece of paper. Electroencephalographers or Electroencephalography technician have decoded and detected patterns along the signals by visually inspecting them \citep{Schomer2010}.   Nowadays clinical EEG still remains a visually interpreted test \citep{Hartman2005}.

In contrast, automatic processing, or quantitative EEG, was based first on analog electronic devices and later on computerized digital processing methods \citep{Jansen1991}.  They implemented mathematically and algorithmically complex procedures to decode the information with good results \citep{Yuste2017}.  The best materialization of the automatic processing of EEG signals rests in the BCI discipline, where around $71.2\%$ is based on noninvasive EEG \citep{Guger2017}.

@@

Hence, the traditional and knowledgeable approach was mainly overshadow particularly in BCI Research, and the waveform of the EEG was replaced by pragmatically sound procedures that were difficult to link to existing clinical EEG knowledge.  We aim to help fix this gap by providing a review of the methods which emphasize the waveform, the shape of the EEG signal and that can help to decode them in a supervised and semi-automated way.

The aim of this study is twofold: first to review current literature of EEG processing techniques which are based on analysis of the waveform.  The second is to evaluate and study these methods by analyzing its classification performance against a pseudo-real dataset. We believe that the importance of waveform analysis methods, as described here, is that by using this methodology, collaboration could be fostered because there is a clear description and characterization of the signal, where the extensive literature which explores clinical EEG can be reviewed from the same shared perspective \citep{Nijboer2009,Wei2017}.

@@

The Electroencephalography consists on the measurement of small variations of electrical currents over the scalp.  It is one of the most widespread used method to capture brain signals and was initially developed by Hans Berger in 1924 and has been extensively used for decades to diagnose neural diseases and other medical conditions.

The first characterization that Dr. Berger detected was the Visual Cortical Alpha Wave, the \textit{Berger Rythm} \citep{Jansen1991}.  He understood that the amplitude and shape of this rhythm was coherently associated to a cognitive action (eyes closing).  
We should ask ourselves if the research advancement that came after that discovery would have happened if it weren't so evident that the shape alteration was due to a very simple and verifiable cognitive process.

The EEG signal is a highly complex multi-channel time-series.  It can be modeled as a linear stochastic process with great similarities to noise \citep{Thakor2004}.  It is measured in microvolts, and those slightly variations are contaminated with heavy endogenous artifacts and exogenous spurious signals.

\begin{figure}[H]
\centering
\includegraphics[height=6cm,width=12cm]{images/eegsample.png}
\caption{Sample EEG signal captured and plotted on a 2D image using the popular and commercial EPOC Emotiv using EEGLAB.}
\label{fig:sampleeeg}
\end{figure}

The device that captures these small variations in current potentials over the scalp is called the Electroencephalograph.  Electrodes are located in predetermined positions over the head, usually embedded in saline solutions to facilitate the electrophysiological interface and are connected to a differential amplifier with a high gain which allowed the measurement of tiny signals. Although initially analog devices were developed and used, nowadays digital versions connected directly to a computer are pervasive.  A detailed explanation on the particularities and modeling of EEG can be obtained from \citep{Jackson2014}, and a description of its electrophysiological aspects from \citep{Haberman2012}.

Overall, EEG signals can be described by their phase, amplitude,  frequency and \textit{waveform}.  The following components regularly  characterize EEG signals:

\begin{itemize}
\item Artifacts:  These are signal sources which are not generated from the CNS, but can be detected from the EEG signal.  These are called endogeneous or physiological when they are generated from a biological source like muscles, ocular movements, etc., and exogeneous or non-physiological when they have an external electromagnetic source like line induced currents or electromagnetic noise\citep{Weeda2012}.
\item Non-Stationarity: the statistical parameters that describe the EEG as a random process are not conserved through time, i.e. its mean and variance, and any other higher-order moments are not time-invariant \citep{Jansen1991}.
\item DC drift and trending: in EEG jargon, which is derived from concepts of electrical amplifiers theory, Direct Current (DC) refers to very low frequency components of the EEG signal which varies around a common center, usually the zero value.  DC drift means that this center value drifts in time.  Although sometimes considered as a nuisance that needs to get rid of, it is known that very important cognitive phenomena like Slow Cortical Potentials or Slow Activity Transients in innates do affect the drift and can be used to understand some particular brain functioning \cite{Schomer2010}.
\item Basal EEG activity: the EEG is the compound summation of myriads of electrical sources from the CNS.  These sources generate a baseline EEG which shows continuous activity with a small or null relation with any concurrent cognitive activity or task.
\item Inter-subject and intra-subject variability: EEG can be affected by the person's behavior like sleep hygiene, caffeine intake, smoking habit or alcohol intake previously to the signal measuring procedure \citep{Farzan2017}.
\end{itemize}

Regarding how the EEG activity can be related to an external stimulus that is affecting the subject, it can be considered as

\begin{itemize}
\item Spontaneous: generally treated as noise or basal EEG.
\item Evoked: activity that is time-locked to an incoming stimulus or an executed motor action.  In contrast it is often called Induced activity
\end{itemize}

\noindent Additionally, according to the existence of a repeated rhythm, the EEG activity can be understood as

\begin{itemize}
\item Rhythmic: EEG activity consisting in waves of approximately constant frequency.  It is often abbreviated RA (regular rythmic activity).
\item Arrhythmic: EEG activity in which no stable rhythms are present.
\item Dysrhythmic: Rhythms and/or patterns of EEG activity that characteristically appear in patient groups or rarely or seen in healthy subjects.
\end{itemize}

The number of electrodes and their positions over the scalp determines a \textbf{Spatial Structure}: signal elements can be generalized, focal or lateralized, depending on in which channel (i.e. electrode) they are found.

%Finally, indexes can be derived as  CFC, Cross Frequency Coupling,  Phase-Amplitude Coupling, Phase-Phase Coupling.

The shape of the signal, the waveform, can be defined as the graphed line that represents the signal's amplitude plotted against time. It can also be called EEG biomarker,  EEG pattern, signal shape, signal form and a morphological signal \citep{Jansen1991}.

The signal context is crucial for waveform characterization, both in a spatial and in a temporal domain \citep{Jansen1991}.  Depending on the context, some specific waveform can be considered as noise while in other cases is precisely the element which has a cognitive functional implication.

A waveform can have a characteristic shape, a rising or falling phase, a pronounced plateau or it may be composed of ripples and wiggles. In order to describe them, they are characterized by its amplitude, the arch, whether they have (non)sinusoidal shape, by the presence of an oscillation or imitating a sawtooth (e.g. Motor Cortical Beta Oscillations).  The characterization by their sharpness is also common, particularly in Epilepsy, and they can also be identified by their resemblance to spikes (e.g. Spike-wave discharge).

Other depictions may include, subjective definitions of sharper, arch comb or wicket shape, rectangular, containing a decay phase or voltage rise, peaks and troughs, short term voltage change around each extrema in the raw trace.  Derived ratios and indexes can be used as well like peak and trough sharpness ratio, symmetry between rise and decay phase and slope ratio (steepness of the rise period to that of the adjacent decay period).  For instance,  wording like "Central trough is sharper and more negative that the adjacent troughs" are common in the literature.

Other regular characterizations which are based on shape features may include:

\begin{itemize}
\item Attenuation: Also called suppression or depression. Reduction of amplitude of EEG activity resulting from decreased voltage. When activity is attenuated by stimulation, it is said to have been "blocked" or to show "blocking".
\item Hypersynchrony. Seen as an increase in voltage and regularity of rhythmic activity, or within the alpha, beta, or theta range. The term implies an increase in the number of neural elements contributing to the rhythm.
\item Paroxysmal. Activity that emerges from background with a rapid onset, reaching (usually) quite high voltage and ending with an abrupt return to lower voltage activity. Though the term does not directly imply abnormality, much abnormal activity is paroxysmal.
\end{itemize}

\begin{itemize}
\item Monomorphic: Distinct EEG activity appearing to be composed of one dominant activity
\item Polymorphic: distinct EEG activity composed of multiple frequencies that combine to form a complex waveform.
\item Transient. An isolated wave or pattern that is distinctly different from background activity.
\end{itemize}

The traditional clinical approach consisted in analyzing the paper strip that was generated by the plot of the signal obtained from the device.  Expert technician and physicians analyzed visually the plots looking for specific patterns that may give a hint of the underlying cognitive process or pathology.   Atlases and guidelines were created in order to help in the recognition of these complex patterns.   Even Video-electroencephalography scalp recordings are routinely used as a diagnostic tools \citep{Giagante2003} .  The clinical EEG research has also focused on temporal waveforms, and a whole branch of electrophenomenology has arisen around EEG \textit{graphoelements} \citep{Schomer2010}.

Sleep Research has been studied in this way by performing Polysomnographic recordings (PSG)  \citep{Rodenbeck2006}, where the different sleep stages are evaluated by visually marking waveforms or graphoelements in long-running electroencephalographic recordings, looking for patterns based on standardized guidelines.   Visual characterization includes the identification or classification of certain waveform components, or transient events, based on a subjective characterization (e.g. positive or negative peak polarity) or the location within the strip.  It is regular to establish an amplitude difference between different waveforms from which a relation between them is established and a structured index are created (e.g. sleep K-Complex is well characterized based on rates between positive vs negative amplitude) (REFERENCE).  Other relevant EEG patterns for sleep stage scoring are alpha, theta, and delta waves,  sleep spindles, polysplindles, K-complexes, vertex sharp waves (VSW), and sawtooth waves (REM Sleep).

Moreover, EEG data acquisition is a key procedure during the assessment of patients with focal Epilepsy for potential seizure surgery, where the source of the seizure activity must be reliably identified. The onset of the Epileptic Seizure is defined as the first electrical change seen in the EEG rhythm which can be visually identified from the context and it is verified against any clinical sign indicating seizure onset.  The interictal epileptiform discharges (IEDs) are visually identified from the paper strip, and they are also named according to their shape: spike, spike and wave or sharp-wave discharges\citep{EEGIntro}.

@@

%Finally, there are specific EEG patterns that can be used to determine Depth of anestisia.

@@

@@

The following criteria is proposed to identify methods which are based on the signal's waveform:

\begin{enumerate}
\item The analysis take account the shape of the plot of the signal.
\item The pattern can be identified and verified by visual inspection.
\item The pattern matching is performed in time-domain.
\item The method encompass a feature extraction procedure.
\item The feature extraction procedure allows to create a template dictionary.
\item A single-channel processing scheme.
\end{enumerate}

As described in (Mixed Domain Signal Analysis) the Pattern Matching problem in Signal processing is finding a signal given the region that best describes the structure of the pattern.   All these algorithm were implemented on MATLAB 2014a (Mathworks Inc., Natick, MA, USA).

\subsection{EEG Waveform Analysis Algorithms}

Shape or waveform analysis methods are considered as nonparametric (in opposition to statistical or dynamical models).  They explore signal's time-domain metrics or even derive more complex indexes from it \citep{Thakor2009}.

One of the earliest approach to automatically process EEG data is the Peak Picking method.  Although of limited usability, peak picking has been used to determine latency of transient events in EEG \citep{Jaskowski2000,Zhang2011}.  Straightforward in its implementation, it consists in selecting a component, a simple component based on the expected location of its more prominent deflection \citep{Ouyang2017}.  Of regular use in ERP Research, the name of many of the EEG features reference directly a peak within the component, e.g. P300 or P3a P3b or N100.  This leads to a natural way to classify them visually by selecting appropriate peaks and matching their positions and amplitudes in an orderly manner.  The letter provides the polarity (Positive or Negative) and the numbering shows the time referencing the stimulus onset, or the ordinal position of each peak (first, second, etc).

A related method is used in \citep{Alvarado-Gonzalez2016} where the area under the curve of the EEG is sumarized to derive a feature.  This was even used in the seminal work of Farwell and Donchin on P300 \citep{Farwell1988,WolpawJonathanR2012}. Additionally, a logarithmic graph of the peak-to-peak amplitude which is called amplitude integrated EEG (aEEG) \citep{Shah2015} is used nowadays in Neonatal Intensive Care Units.

Other works on EEG explored the idea to extend human capacities analyzing EEG waveforms \citep{Klein1976} where a feature from the amplitude and frequency of its signal and its derivative in time-domain is used.  Moreover, other works explored the use of Mathematical Morphology \citep{Yamaguchi2009}, where the time-domain structure of contractions and dilations were studied, and finally the seminal proposals of Fujimori, Uchida and the PAA algorithm as few of the earliest proposals where the idea of capturing the shape of the signal were established.

%Althought CASENET REFERENCe explored "waveform" structure they were purely based on spike detection based on feeding artificial neural networks.

 %(fujimori, PAA, etc).  Uchida 1999

%\subsubsection{Averaging Methods}
%
%The methods that allow to identified waveforms are used to determine different alignments while averaging epochs.
%
%This long-standing problem has been tackled from different perspectives.  Woody's Template Matching is perhaps the best effort as well as Pham, and Tuan and others EML.
%
%Dictionary, template based method.  Many do not consider the waveform, albeit they do use a set of templates obtained from a dictionary.
%
%Cross Covariance with Template or Cross Correlation or ACF: Autocorrelation Function: PHAM METHOD
%
%Applying a FIR filter with templates 
%
%Krusienski et al 2007
%Serby et al 2005
%
%extended to wavelent analysis.
%
%Dynamic Time Warping and other Warp Averaging methods were also used.
%
%Wrapping Analysis
%
%Warp averaging

@@

\begin{itemize}
\item Matching Pursuit
\item Permutation Entropy
\item Slope Horizontal Chain Code
\item Histogram of Gradient Orientations
 \item Merging of Increasing and Decreasing Sequences
\item Local Binary Patterns (1-D LNBP, 1D-LBP and LBP)
\end{itemize}

All these methods provide a feature that can be used as a template, whereas all of them are based on metrics that can be extracted from the shape of the signal.  These features can be used to create dictionaries or templates databases.  These templates provide the basis for the pattern matching algorithm and offline classification.

\subsection{Matching Pursuit - MP 1 and MP 2}

Pursuit algorithms refer, in their many variants, as blind source separation(REF) techniques that assume that the EEG signal is a linear combination of different sources that comes from template's dictionaries.  The underlying idea is to find the best template out of a dictionary that matches against the signal.  The algorithm works iteratively, starting from the raw signal and then, the template is substracted from the signal in Graham-Schmidt approach and the process iterates. (ref Cohen book, Sanei book, Mallat and Zhang 1993).

On the other hand, Morphological Component Analysis is a variant form of Blind Source Separation which can be considered the extension of Matching Pursuit algorithm. (cita a esa parte del paper).

\subsection{Permutation Entropy - PE}

Bond and Pompe Permutation Entropy has been extensively used in EEG processing, particularly for Epilepsy pre-ictal detection.  This method generates a code based on the orderly arrangement of sequential samples, and then derives a metric which is based on the amount of entropy of each code within the signal.   For instance, a pure random signal, will achieve the maximum entropy value due to the probability of each code being equal for all of them. This method is the best representative of Waveform Complexity: Lempel-Ziv method L-Z complexity are other variants which use a different definition (Permutation Entropy a new feature for brain computer interfaces).

\subsection{Slope Horizontal Chain Code - SHCC}

This method is an extension of Slope Chain Code (SCC) (ref a SCC paper y a SHCC paper), and a code of a sequence of sample points is generated based on the angle between the horizontal segment and any segment, one by one.  This method can be encompass as a syntactic analysis technique, where the EEG segment is represented as a series of elementary patterns, similar to tokens, due to the quantization of angles.

\subsection{Histogram of Gradients Orientations - HGO}

This Histogram of Gradient Orientations is based on using Computer Vision techniques to extract directly information from a signal based on the plot of a signal on a 2D image.  This mimicks exactly what a person is doing by visually inspecting the plot, and the waveform.  To do that, the region of the image where the signal is located is divided in a $4 x 4$ block and the signal bidimensional gradients is calculated.  For each block within the patch, a histogram of gradient orientations, 8 circular orientations, is calculated.  This histogram is concatenated for all the 16 blocks and a feature is thus formed.  The details of the method can be found here \citep{Ramele2016}.  This method was implemented using the VLFeat  \citep{Vedaldi2010} Computer Vision libraries.

%\subsection{Merger of Increasing and Decreasing Sequences}

%Stepwise downsampling

@@

Results are shown in Table \ref{tab:results} and in Figure \ref{fig:performancetest},\ref{fig:performancetestlatency} and \ref{fig:performancetestamplitude}.  Table \ref{section:results} shows the single-trial performance while identifying each letter of the standard P300 Speller Matrix, and the channel where the best performance was obtained.   Figure \ref{fig:performancetest} shows the performance curves for eight algorithms.  Each one represents the percentage of letters that were actually predicted by the combination of feature plus a classification method.  The data were divided in two, and the first 15 letters were used to derive the dictionary of templates $T$ while the remaining 20 letters were used to measure the classification accuracy.   Figure \ref{fig:performancetestlatency} shows the same results obtained when a random uniform latency shift was added to each trial.   Finally, Figure \ref{fig:performancetestamplitude} represents the performance values obtained when the amplitude of the N1 and P3 component of the template are randomly reduced (uniform distribution).
